GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-10-04 figs/GSE158422_sensitivity.Rmd
moving on to sensitivity analysis

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

Functional class sorting is widely used for pathway analysis using tools like GSEA,
yet there is no consensus in how this can be conducted for Illumina Infinium
methylation array data.

Here we propose a simple approach which involves the following:

1. Limma test on probes.

2. For each gene, calculate the median t-statistic from step 1.

3. Use this median t-stat in Camera pre-ranked test for gene sets.

In this example, I'm using matched infinium EPIC 850k data from (n=37) normal and lung cancer samples (GSE158422).
The data was previously preprocessed and normalised using the minfi package (see the folder called "misc").

Here the gene sets are obtained from Reactome.

@@

This analysis was run on a 8C/16T computer with 64GB RAM running at 3.8 GHz.
This workflow used 34 GB RAM and took
RAM usage can be moderated by reducing the parallel core count.

@@

# GMEA: all probes

There are two obvious ways to perform GMEA:

1. limma-agg-1stt (la1).
In this approach, the limma test is conducted on probes, followed by aggregation of the differential
methylation t-scores. Each gene gets a median t-stat value which is evaluated downstream by a 1 sample
t-test.

2. agg-limma-1stt (al1).
Here, the probe beta values for each gene are aggregated to a median and then limma is run on these
values.
To determine gene set enrichment, a 1-sample t-test is conducted.

## LA: limma aggregate 1-sample t-test all probes

```{r,probelimma1}

@@

fit.reduced <- lmFit(mval,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=4, number = Inf)
dm <- merge(myann,dm,by=0)
dm <- dm[order(dm$P.Value),]
rownames(dm) <- dm$Row.names
dm$Row.names=NULL

hist(dm$t,breaks=30,main="all probes")
hist(subset(dm,Relation_to_Island=="Island")$t,main="CpG island")
hist(subset(dm,Relation_to_Island=="N_Shore")$t,main="CpG shore")
hist(subset(dm,Relation_to_Island=="N_Shelf")$t,main="CpG shelf")

hist(subset(dm,Regulatory_Feature_Group=="",main="No genomic feature")
hist(subset(dm,Regulatory_Feature_Group=="Promoter_Associated")$t,main="promoter")
hist(subset(dm,Regulatory_Feature_Group=="Gene_Associated")$t,main="gene")
hist(subset(dm,Regulatory_Feature_Group=="Unclassified")$t,main="unclassified")

head(dm) %>%
  kbl(caption="Top DM probes") %>%
  kable_styling("hover",full_width=FALSE)

agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    if ( length(tstats) > 2 ) {
      ttest <- t.test(tstats)
      pval <- ttest$p.value
    } else {
      pval = 1
    }
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
      "median"=mymedian, pval=pval)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg <- agg(dm,cores=CORES)
toc()

head(dmagg$gme_res_df,20) %>%
  kbl(caption="top gmea genes (t-test)") %>%
  kable_paper("hover", full_width = F)

@@

Check whether the distribution is even - it's not.

```{r,tmeddist}

hist(dmagg$gme_res_df$median,breaks=30) ; grid()

hist(dmagg$gme_res_df$mean,breaks=30) ; grid()

@@

1-sample t-test for gene set enrichment.

```{r,t_enrichment_test1}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

# try mean and median
m <- dmagg$gme_res_df[,"mean",drop=FALSE]
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(lares$t_mean)

# need to get the median column before analysis
m <- dmagg$gme_res_df[,"median",drop=FALSE]
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(lares$t_median)

head(lares,30) %>%
  kbl(caption = "Top significant genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

head(lares[order(-abs(lares$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

lasig <- subset(lares,fdr<0.05)
nrow(lasig)

head(lasig[order(-abs(lasig$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets after FDR filtering using LA approach") %>%
  kable_paper("hover", full_width = F)

plot(lares$t_median,-log10(lares$pval),pch=19,col="gray")
grid()
points(lasig$t_median,-log10(lasig$pval),pch=19,col="red")

## AL: Aggregate limma t-test all probes

Determine the median of probe values

```{r,med1}

mx <- mval

gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$gnl <- gnl

# compare median and mean based aggregation
mymed <- function(g) {
  probes <- rownames(myann[grep(g,myann$gnl),])
  rows <- which(rownames(mx) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mx[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

med <- mclapply(gn,mymed,mc.cores=CORES)
med <- med[lapply(med,length)>0]
medf <- do.call(rbind,med)

boxplot(list("normal"=matrix(colMeans(mval),ncol=2)[,2],"tumor"=matrix(colMeans(mval),ncol=2)[,1]),
  main="mean mval for control and tumor samples (all probes)")

boxplot(list("normal"=matrix(colMeans(medf),ncol=2)[,2],"tumor"=matrix(colMeans(medf),ncol=2)[,1]),
  main="mean mval for control and tumor samples (gene associated)")

Then do a limma test on median gene methylation.
See if it is more robust than other approaches.

```{r,limma1}

fit.reduced <- lmFit(medf,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
magg <- topTable(fit.reduced,coef=ncol(design), number = Inf)

hist(magg$t)

head(magg, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)

Very curious that the distribution is negatively skewed now.

Let's do another test to see whether this is correct or not.

```{r,test}

mytest <- lapply(1:nrow(medf),function(i) {
  ctl <- matrix(medf[i,],ncol=2)[,2]
  tum <- matrix(medf[i,],ncol=2)[,1]
  tt <- t.test(tum,ctl)
  t=tt$statistic ; p=tt$p.value ; est=rev(tt$estimate)
  names(est) = c("ctl","tum")
  c(t,p,est)
} )

mytest <- as.data.frame(do.call(rbind,mytest))
colnames(mytest) = c("t","p","ctl","tum")
rownames(mytest) <- rownames(medf)
hist(mytest[,"t"])

mytest <- mytest[order(mytest$p),]
head(mytest,20) %>%
  kbl(caption="Top significant differential gene methylation results using dumb test") %>%
  kable_paper("hover", full_width = F)

nrow(subset(mytest,p<0.05 & t>0))
nrow(subset(mytest,p<0.05 & t<0))

Hypomethylated genes still outnumber hypermethylated genes.

It does skew downwards after aggregation.

1-sample t-test for enrichment

```{r,ttest}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      pval = wt$p.value
      res <- c(name,myn,mymean,mymedian,pval)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

m <- as.data.frame(magg$t)
rownames(m) <- rownames(magg)
colnames(m) <- "t"

alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)

hist(alres$t_mean)

@@

Check whether the distribution is even - it's not.

```{r,laphist}

hist(dmpagg$gme_res_df$median,breaks=30) ; grid()

1-sample t-test for gene set enrichment.

```{r,t_enrichment_test1_promoter}

# need to get the median column before analysis
m <- dmpagg$gme_res_df[,"median",drop=FALSE]

tic()
lapres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
toc()

head(lapres,30) %>%
  kbl(caption = "Top significant genesets using LA approach for promoters") %>%
  kable_paper("hover", full_width = F)

head(lapres[order(-abs(lapres$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

lapsig <- subset(lapres,fdr<0.05)
nrow(lapsig)

head(lapsig[order(-abs(lapsig$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets after FDR filtering using LA approach for promoters") %>%
  kable_paper("hover", full_width = F)

plot(lapres$t_median,-log10(lapres$pval),pch=19,col="gray")
grid()
points(lapsig$t_median,-log10(lapsig$pval),pch=19,col="red")

## AL: Aggregate limma t-test Promoter only

Determine the median of probe values.

```{r,med1p}

gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$gnl <- gnl
myannp <- myann[grep("Promot",myann$Regulatory_Feature_Group),]

mxp <- mval[rownames(mval) %in% rownames(myannp),]

mymed <- function(g) {
  probes <- rownames(myannp[grep(g,myannp$gnl),])
  rows <- which(rownames(mxp) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mxp[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

medp <- mclapply(gn,mymed,mc.cores=CORES)
medp <- medp[lapply(medp,length)>0]
medfp <- do.call(rbind,medp)

Then do a limma test on median gene methylation.
See if it is more robust than other approaches.

```{r,limma1p}

fit.reduced <- lmFit(medfp,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
maggp <- topTable(fit.reduced,coef=ncol(design), number = Inf)

head(maggp, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)

1-sample t-test for enrichment promoter.

```{r,ttestp}

m <- as.data.frame(maggp$t)
rownames(m) <- rownames(maggp)
colnames(m) <- "t"

alpres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)

alpsig <- subset(alpres,fdr<0.05)

nrow(alpsig)

nrow(subset(alpsig,t_median>0))

head(alpsig[order(-alpsig$t_median),],20) %>%
  kbl(caption="hypermethylated pathways") %>%
  kable_paper("hover", full_width = F)

nrow(subset(alpsig,t_median<0))

head(alpsig[order(alpsig$t_median),],20) %>%
  kbl(caption="hypomethylated pathways") %>%
  kable_paper("hover", full_width = F)

## Promoter Overlap

```{r,ol1p}

lapsig_up <- rownames(subset(lapsig,t_median>0))
lapsig_dn <- rownames(subset(lapsig,t_median<0))

alpsig_up <- rownames(subset(alpsig,t_median>0))
alpsig_dn <- rownames(subset(alpsig,t_median<0))

v1 <- list("LA up"=lapsig_up, "LA dn"=lapsig_dn,
  "AL up"=alpsig_up,"AL dn"=alpsig_dn)

>>>>

---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

In the previous analysis, it was shown that the "AL" (aggregation-limma-enrichment) method
yielded good performance with a cohort of cancer-normal tissue pairs.

In this piece of work, we want to ascertain the sensitivity of this approach as compared to
existing methods such as gsameth and ebGSEA.

To achieve this we will analyse with the full 37 patient sample pairs, and repeatedly downsample
the dataset down to n=3 patients.
What we expect is that the "AL" method could have a higher degree of recall as compared to these
other approaches.
We expect that "AL" could identify gene sets with larger effect sizes (enrichment scores) using
fewer samples.
This means that for a given set of datasets, this approach could yield more findings, or
alternatively if pathway analysis is the primary outcome of a study, it need not be as large which
can save money and difficulty with establishing large sample collections.

@@

This analysis was run on a 16C/32T computer with 128GB RAM running at 4.2 GHz.
RAM usage can be moderated by reducing the parallel core count.

@@

# AL GMEA: all probes

Start with curating the sample info.

```{r,samplesheet}

@@

@@

## Functions

```{r,functions}

# magg is a general function that aggregates probe methylation measurements by gene
# in this variation it aggregates by MEAN, which performs better than median
magg <- function(mval,myann,cores=1){
  gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  myann$gnl <- gnl
  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
  mx <- mval[rownames(mval) %in% keep,]
  mymed <- function(g) {
    probes <- rownames(myann[grep(g,myann$gnl),])
    rows <- which(rownames(mx) %in% probes)
    if ( length(rows) > 1 ) {
      b <- mx[rows,]
      med <- apply(b,2,mean)
      med <- matrix(med,nrow=1)
      colnames(med) <- colnames(b)
      rownames(med) <- g
      return(med)
    }
  }
  med <- mclapply(gn,mymed,mc.cores=cores)
  med <- med[lapply(med,length)>0]
  medf <- do.call(rbind,med)
  return(medf)
}

# chragg is a specialised function that performs the magg routine on chromosomes in parallel
chragg <- function(mval,myann,cores=1){
  annodf <- as.data.frame(anno)
  keep <- rownames(subset(myann,UCSC_RefGene_Name!=""))
  mx <- mval[rownames(mval) %in% keep,]
  chrs <- unique(anno$chr)
  myorder <- unlist(lapply(chrs,function(mychr) { nrow( annodf[annodf$chr==mychr,] ) } ))
  chrs <- chrs[order(-myorder)]
  leadercores <- floor(sqrt(cores))
  workercores <- ceiling(sqrt(cores))
  chrmedf <- mclapply(chrs,function(chr) {
    chrfrag <- annodf[annodf$chr==chr,]
    chrprobes <-rownames(chrfrag)
    chrmx <- mx[rownames(mx) %in% chrprobes,]
    chranno <- myann[rownames(myann) %in% chrprobes,]
    chrmedf <- magg(mval=chrmx,myann=chranno,cores=workercores)
    return(chrmedf)
  },mc.cores=leadercores)
  medf <- do.call(rbind, chrmedf)
  return(medf)
}

# agglimma is a helper function that runs limma analysis on the aggregated methylation data
# and fixes any duplicated gene names - most significant one is retained
agglimma <- function(medf,design) {
  fit.reduced <- lmFit(medf,design)
  fit.reduced <- eBayes(fit.reduced)
  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
  nondup <- !duplicated(dmagg$ID)
  dmagg <- dmagg[nondup,]
  rownames(dmagg) <- dmagg$ID
  dmagg$ID = NULL
  return(dmagg)
}

# ttenrich is a function designed to conduct analysis of differential methylation.
# It takes the t-statistics from limma for a geneset and determines whether this gene
# set has a distribution different to the null (zero) using a one-sample two-sided t-test.
ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","pval")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$pval),]
  res_df$logp <- -log10(res_df$pval )
  res_df$fdr <- p.adjust(res_df$pval,method="fdr")
  res_df[order(abs(res_df$pval)),]
  return(res_df)
}

@@

Run AL method on methylation data

```{}

# aggregation step
medf <- chragg(mval,myann,cores=CORES)

# differential analysis
dmagg <- agglimma(medf,design)
hist(dmagg$t)

head(dmagg, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)
# get t-stats
m <- as.data.frame(dmagg$t)
rownames(m1) <- rownames(magg1)
colnames(m1) <- "t"

# enrichment test
alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(alres$t_mean)

@@
